Reformulating phylogenetic mixed models to improve flexibility and speed

### 2019-06-04 phyloglmm_ms.tex

However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.

@@

Classic PCMs usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables may vary according to the phylogenetic relationship across species as well.
Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using a mixed-effect model. 
It is plausible that taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic signals in different ways and looked at species response to phylogenetic signals with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table \ref{table:model} provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
\multirow{2}{*}{
\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
\end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez} \pkg{phyr}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{List of phylogenetic generalized linear models and R packages.}
\label{table:model}
\end{table}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
Our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic signal from both predictors and tips/species, as well as residual variation.
As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% FIXME: move to methods/simulations
% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

CONT HERE

When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random intercepts are more approriate than random slopes, or including both of them in the model.
It may be easier to be conservative to include both of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model. 
Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes? and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}? 
% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.

\subsection*{Process and observation variability and confounding}

% \bmb{explain: classical scenario for PIC is when we have species-level data (i.e., species averages or ``typical'' values).  What do we do when we have multiple observations per species?  Why can't we just average? (cf. \cite{murtaugh2007simplicity}) with a nested, balanced design, averaging by group and doing 1-way ANOVA gets the same answer for fixed effects as LMMs). \textbf{answers}: (1) unbalanced data (can handle by weighting by $n$, equivalent to assuming homoscedasticity and using inverse-variance weights); (2) non-Gaussian responses (averaging doesn't make sense) [can we use offsets to handle this case? don't know]; (3) maybe for equivalent of randomized block designs? (4) when within-species variance is actually of interest}
Process and observation errors are highly confounded in phylogenetic regression if it is not handled correctly. 
For single measurement per species model, any method that can account for at least two sources of variation (one for the residual variance, and at least one for the process variation) will be sufficient to a first approximation (i.e. Pagel's $\lambda$, Blomberg's $K$).
If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic process by substituting tip variation with residual variation, which is equilvent to Felsenstein's ``worst case".
There are a few ways to deal with this case.
For balanced data one can summarize multiple observations to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
For unbalanced data, one can handle by weighting by number of observations per species, which is equivalent to assuming homoscedasticity using inverse-variance weights. 
Alternatively, when within-species variance is actually of interest, accounting for within-species variance (i.e. adding species-level random effect in our example) can automatically handle multiple observations per species.

@@

Our analysis covers the classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models).
Even within the scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-BM evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \cite{butler2004phylogenetic}), Bayesian approaches \cite{hadfield2010general}.
More broadly, the simple independence error approach we developed here offers a more efficient and mathematical equivalent way to do Brownian motion evolution process phylogenetic comparative analysis. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistically models, users should be aware the amount of data they have and the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed} or possibility of Bayesian approaches \cite{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} allows users to fit phylogenetic mixed models to the fullest ( that other frequentist platforms lack of (i.e. large data, unbalanced species observations, complex random effects) and explore new ideas.

@@

We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
First, this new approach is magnitudes faster compared to existing phylogenetic mixed models without losing robustness/accuracy and can easily combined with any statistically mixed model framework. 
Last and most importantly, it is more flexible in fitting large phylogenies, large volume of data, unbalanced data sets, and complex random effects such as slope correlations.

>>>>

However, if each species' traits are observed more than once, possibly under different conditions, we can potentially distinguish a third level of variation; in this case, phylogenetic variation and tip variation are now both part of the evolutionary process error (which we will call tip variation as intercept-level variation) while the among species residual variation is associated with among observation variation within each species.

@@

Classic PCMs usually examines the response (trait or distribution) to vary according to the phylogenetic relationship across species, but it is plausible the changes in effect (or levels) of the predictor variables may vary according to the phylogenetic relationship across species as well.
Suppose we have examined a collection of species that came from two groups, and wish to know whether their brain size (Y) is proportional to their body size (X) (Felsenstein's example \citep{felsenstein1985phylogenies}) using a mixed-effect model. 
It is plausible that taxonomic groups with similar body sizes may vary in brain size (random intercepts), or taxonomic groups may vary in the relationship between brain and body size (random slopes). 
%\bmb{maybe we can focus on random slopes here? Need to point out that example studies below are \emph{exceptions} to the general assumption of random intercepts only (how did they manage to do their inference?)}
Several recent studies have incorporated phylogenetic variations in different ways and looked at species response to phylogenetic variations with changes in environmental factors.
For example, \cite{nowakowski2018phylogenetic} considered phylogenetically correlated slopes in response to habitat conversion when studying the abundance of amphibian species using a Bayesian implementation of phylogenetic GLMM, while \cite{li2017canfun} considered phylogenetically correlated species nested within sites for plant abundance via phylogenetic GLMM approach. 
The tools available for extending phylogenetic relationships to predictor variables (or predictor level variation) are relatively inflexible; thus, computationally sophisticated biologists have turned to more flexible Bayesian approaches, despite their computational burden \citep{hadfield2010mcmc, burkner2016brms}.
Table \ref{table:model} provides a summary of model complexities, platforms and data constraints for phylogenetic comparative analysis.

@@

%% FIXME: chart looks ugly
\begin{table}[]
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Model} & \textbf{Method} & \textbf{Data} & \textbf{Platform} \\ \hline
\multirow{2}{*}{
\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Model (GLM)
\end{tabular}} & Correlated residual & Single observation & \begin{tabular}[c]{@{}l@{}}\pkg{nlme}:gls, \\ \pkg{ape}:pic\end{tabular} \\ \cline{2-4} 
& \begin{tabular}[c]{@{}l@{}}Residual \\ + phylo intercept\end{tabular} & Single observation & \begin{tabular}[c]{@{}l@{}}Pagel's $\lambda$\\ Blomberg's $k$ \\ via \pkg{nlme}:gls\\ \pkg{phylolm}\end{tabular} \\ \hline
\multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Generalized Linear \\ Mixed Model (GLMM)\end{tabular}} & \multirow{2}{*}{Random effect}                                        & \begin{tabular}[c]{@{}l@{}}Single observation\\ Balanced design\end{tabular} & \pkg{pez}, \pkg{phyr}                                                                                            \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{lme4}, \pkg{glmmTMB}                                                                                  \\ \hline
\multirow{2}{*}{Bayesian GLMM}                                                                    & \multirow{2}{*}{Random effect}                                        & Balanced design                                                              & \pkg{MCMCglmm}                                                                                       \\ \cline{3-4} 
                                                                                                  &                                                                       & Unrestricted                                                                 & \pkg{brms}                                                                                           \\ \hline
\end{tabular}
\caption{List of phylogenetic generalized linear models and R packages.}
\label{table:model}
\end{table}

In this paper, we will propose an alternative formulation of phylogenetic generalized linear mixed models that is mathematically equivalent to previous approaches, but more flexible.
Our new formulation can be incorporated in any framework that allows for random effects (such as random intercepts, slopes, and interactions) without the need to implement special correlation structures, by incorporating phylogenetic structures as part of the mean model. 
%% can be incorporated relatively easily into any mixed-model platform that allows new random-effects design matrices to be specified, and can be easily extended along with the features of mixed model such as multiple observational designs, phylogenetic random-slopes models, and multiple (nested or crossed) random effects.
We will compare our technique coded in R packages \pkg{lme4} and \pkg{glmmTMB} with existing R packages (i.e. \pkg{nlme} \citep{pinheiro2014r}, \pkg{phylolm} \citep{ho2014phylolm}, \pkg{pez} \citep{pearse2015pez} \pkg{phyr} (Li et al, unpublished), \pkg{MCMCglmm} \citep{hadfield2010mcmc}, and \pkg{brms} \citep{burkner2016brms}) fitting models to data from simulated model that incorporates phylogenetic variation from both predictors and tips/species, as well as residual variation.
As well as being flexible, we also expect that our approach will be fast, because \pkg{lme4} and \pkg{glmmTMB} use efficient approaches. 
%% FIXME: move to methods/simulations
% In principle, for any given valid method that matches the simulation model should eventually converge to the simulated parameters. 
% We end by discussing opportunities and practicalities of our method and comments on the state of art for this area of research.

@@

When analyzing relationships among species between traits, it is entirely plausible that change in effect of predictor variables have on traits are changing with respect to different grouping of species or even in phylogenetically correlated way (phylogenetic random slopes).
Nevertheless, it is hard to account for all forms of complexities and decide if phylogenetic random effects are more approriate than simple grouping random effects, or including all of them in the model.
However, the model complexity depends on the experimental design and whether there are enough data to estimate these different levels of variation.
Some of these variations can be highly confounded in phylogenetic regression if it is not handled correctly or if there are not enough data.
For example, for experimental designs with single measurement per species, any method that can account for at least two sources of variation, such as Pagel's $\lambda$ (where one parameter quantifying phylogenetic variation and one for the residual variation) will be sufficient to a first approximation.
If multiple observations are allowed per species, then using methods such as Pagel's $\lambda$ can potentially incorrectly estimate the amount of phylogenetic variation by confounding tip variation with residual variation, which is equivalent to Felsenstein's ``worst case".
There are a few ways to deal with this case.
For balanced data one can summarize multiple observations to a single measurement (for example, mean or median) per species to avoid confounding residual and tip variations.
For unbalanced data, one can handle by weighting by number of observations per species, which is equivalent to assuming homoscedasticity using inverse-variance weights. 
Alternatively, when within-species variance is actually of interest, accounting for within-species variance (i.e. adding species-level random effect in our example) can automatically handle multiple observations per species.

It may be easier to be conservative to include both (phylogenetic random effects and grouping random effects) of them but simplify the phylogenetic relationships (at the random slopes level) and think about the random-slopes model in a strictly hierarchical setting (i.e., estimating different slopes for each family, or taxon \citep{bunnefeld2012island}) - the PGLMM collapses to a standard random-slopes model. 
Users should be aware of two important questions when fitting random-slope models: How much data do we need in order to practically estimate the random slopes? and Are we making a mistake by ignoring random slopes \citep{schielzeth2008conclusions}? 
% However, in the PGLMM context, as long as there's variation in the predictor among tips, there will be variation among taxa at some level, so random-slopes models will (almost always) be \emph{theoretically} identifiable.

@@

We have presented the range of classical phylogenetic comparative methods (i.e. phylogenetic least squares, linear and mixed models) to fit data that incorporates different levels of phylogenetic variation.
Even within this scope there is additional room for exploration we neglected, such as exploring phylogenetic multivariate response models, non-Brownian evolution processes (such as Ornstein-Uhlenbeck (OU) process which accounts for both selection and drift processes \citep{butler2004phylogenetic}), and Bayesian approaches \citep{hadfield2010general}.
However, the simple approach we developed here offers a more efficient and mathematical equivalent way to handle phylogenetic comparative analysis for a wide range of univariate, Brownian-motion evolutionary models. 
This approach can in principle be combined flexibly with the state of the art phylogenetic mixed models using any platforms that supports independent latent variables such as \pkg{MCMCglmm} \citep{hadfield2010mcmc}; a widely used Bayesian approach to PGLMM.
However, in principle just like GLMM and most statistical models, users should be aware of the constraints of the amount of data they have relative to the complexity of the model they want to fit (i.e. What should we do when we don't have enough data? Should we use a complex model and overfit \citep{barr2013random}, the right balance of complexity and data \citep{baayen2008mixed}, singular fits or possibility of Bayesian approaches \citep{hadfield2010mcmc}).
More importantly, this implementation in \pkg{lme4} and \pkg{glmmTMB} allows users to fit phylogenetic mixed models to the fullest (large data, unbalanced species observations, complex random effects) and explore new ideas.

@@

We have presented a simple approach to fit phylogenetic mixed models that is both more efficient and statistically equivalent approach and comparison of classical PCM to simulated data. 
Our approach is orders of magnitud faster than existing phylogenetic mixed models and it can easily combined with any statistical mixed model framework. 
It is more flexible in fitting large phylogenies, large volumes of data, unbalanced data sets, and complex random effects such as slope correlations.
